Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-03 Rversion/reviewR.qmd
Propagated changes to R version

---
title: "Measuring contrast processing in the visual system using SSVEP"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format: 
  pdf:
    keep-tex: true
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
  output: false
  include: false
---
Department of Psychology and York Biomedical Research Institute, University of York, UK

## Review for Vis Neurosci

@@

pdf('Figures/centresurround.pdf',width=12,height=6.5)

@@

pdf('Figures/linearsystem.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

@@

pdf('Figures/transducer.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

pdf('Figures/twoinputs.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

pdf('Figures/sweepplot.pdf',width=6,height=6)

@@

pdf('Figures/clinicalplot.pdf',width=12,height=12)
par(mfrow=c(2,2))

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2005; @Di-Russo2007]. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@McKeefry1996; @Di-Russo2001]   (see also [@Baseler1997]).

@@

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaper).

![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](Figures/centresurround.pdf){#fig-centresurround}

@@

@@

To  understand these processes better, we will show how sinusoidal input signals might be processed by the visual system to produce SSVEPs. @fig-linear illustrates how sine waves of different contrasts are processed in a linear system. The first panel shows the input sine wave, which would be used to modulate stimulus amplitude over time. Notice that there are five peaks in the waveform during the one second sample, so the stimulation frequency is 5Hz (F1). The second panel shows Fourier transform of the waveform, which contains a substantial peak at this frequency. If we  change the stimulus contrast (i.e. the amplitude of the waveform), the amplitude of the F1 component increases linearly with contrast (right panel).

![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](Figures/linearsystem.pdf){#fig-linear}

Next we can consider the impact of a nonlinearity in processing on the responses. One of the simplest nonlinearities is the function that describes a cell's response to different levels of contrast. In @fig-transducer this is modeled by a hyperbolic ratio function resulting in a saturating non-linearity:

@@

This nonlinearity has clear effects on signal transduction. The sinusoidal waveform is distorted by the nonlinearity, and appears frequency doubled (left panel of @fig-transducer; note that other nonlinearities such as rectification and squaring can have similar effects). The distortion is reflected in the Fourier spectrum (middle panel of @fig-transducer), which now includes responses at integer multiples (known as harmonics) of the stimulation frequency (10, 20 and 30Hz), but no response at the fundamental (5Hz). Finally, the contrast response function is now nonlinear (right panel of @fig-transducer). Although the hyperbolic ratio function (@eq-nakarushton) is monotonic, the CRF resulting from measuring the amplitude of the second harmonic (2F) component contains a slight roll-off at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases, and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler1985; @Peirce2007].

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the squaring operator, and now features energy at the second harmonic (10Hz) and its multiple (middle panel). The contrast response function at the second harmonic is now nonlinear, and saturates at high contrasts (right panel).](Figures/transducer.pdf){#fig-transducer}

To illustrate the effect of contrast gain control [@Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel).](Figures/twoinputs.pdf){#fig-twoinputs}

@@

where $C_\mathrm{mask}$ reflects the contrast of the mask component at a distinct frequency from that of the target. The effect of this extra term is to reduce the target response (see right panel of @fig-twoinputs). For a linear contrast axis, the contrast response function becomes shallower, whereas on a logarithmic contrast axis it maintains its steepness and shifts to the right. Suppressive effects of this kind have been obtained using SSVEP with a variety of different types of mask, including orthogonal overlaid masks, surround masks, and dichoptic masks [@Burr1987; @Ross1991; @Candy2001; @Busse2009; @Salelkar2020; @Cunningham2017; for a meta-analysis see @Baker2021].

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. For example, @Tsai2012 demonstrated that a gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be no studies measuring the full SSVEP contrast response function before and after an extended period of adaptation.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis (see @fig-sweepvep). This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the regression fit to the lowest four data points, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/sweepplot.pdf){#fig-sweepvep}

>>>>

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format: 
  pdf:
    keep-tex: true
bibliography: ../WadeBaker_Review.bib
csl: ../visual-neuroscience.csl
execute:
  echo: false
  output: false
  include: false
---

Department of Psychology and York Biomedical Research Institute, University of York, UK

### Funding statement:
This work was supported by BBSRC grant BB/V007580/1 to DHB and ARW

@@

pdf('Figures/1_centresurround.pdf',width=12,height=6.5)

@@

pdf('Figures/2_linearsystem.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

```{r fullwaverectification}

    # We are about to model response to contrast. Different cell populations detect positive and negative-going contrast and we can model their combined response by performing full-wave rectification as follows: 
  fullWaveRectifiedSignal <- abs(inputModulation)

  # This doubles the frequencies of the resulting FT and also introduces harmonics at higher frequencies but it does not alter the general shape of the contrast vs response function.

  # Now compute the FT on the signal after it has passed through full wave rectification
  FT_transducedModulation <- fft(fullWaveRectifiedSignal) / length(fullWaveRectifiedSignal)

  transducedAmplitude <- abs(FT_transducedModulation[2*inputFrequency_Hz+1])
  transducedSignal[contrastIndex] <- transducedAmplitude

}

pdf('Figures/3_fullwaverectification.pdf',width=12,height=4)
par(mfrow=c(1,3))

  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,0.35))   
  axis(1, at=seq(0,1,0.5), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
  axis(2, at=seq(0,3,1)/10, tck=0.01, lab=F, lwd=2)
  mtext(text = seq(0,1,0.5), side = 1, at=seq(0,1,0.5), line=0.2)     # add the tick labels
  mtext(text = seq(0,3,1)/10, side = 2, at=seq(0,3,1)/10, line=0.2, las=1)  
title(xlab="Time (s)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)    
title(main="Rectified signal", cex.main=1.5)

lines(support_S, fullWaveRectifiedSignal, lwd=3)

# Plot the absolute values of the FFT in the second axis as a bar plot
  plot(frequencySupport[2:length(frequencySupport)]-0.5,abs(FT_transducedModulation[2:length(frequencySupport)]),type='s',lwd=3,axes=FALSE, ann=FALSE, xlim=c(0,30), ylim=c(0,0.08))
  axis(1, at=seq(0,30,10), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
  axis(2, at=seq(0,0.08,0.02), tck=0.01, lab=F, lwd=2)
  mtext(text = seq(0,30,10), side = 1, at=seq(0,30,10), line=0.2)     # add the tick labels
  mtext(text = seq(0,0.08,0.02), side = 2, at=seq(0,0.08,0.02), line=0.2, las=1)  
title(xlab="Frequency (Hz)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)    
title(ylab="Magnitude (a.u.)", col.lab=rgb(0,0,0), line=2.6, cex.lab=1.5)
title(main="Fourier spectrum", cex.main=1.5)

  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,32), ylim=c(0,0.08))   
  axis(1, at=seq(0,32,4), tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
  axis(2, at=seq(0,0.08,0.02), tck=0.01, lab=F, lwd=2)
  mtext(text = seq(0,32,4), side = 1, at=seq(0,32,4), line=0.2)     # add the tick labels
  mtext(text = seq(0,0.08,0.02), side = 2, at=seq(0,0.08,0.02), line=0.2, las=1)  
title(xlab="Contrast (%)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)    
title(ylab="Amplitude (a.u.)", col.lab=rgb(0,0,0), line=2.6, cex.lab=1.5)
title(main="Contrast response function", cex.main=1.5)

@@

pdf('Figures/4_transducer.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

pdf('Figures/5_twoinputs.pdf',width=12,height=4)
par(mfrow=c(1,3))

@@

pdf('Figures/6_sweepplot.pdf',width=6,height=6)

@@

pdf('Figures/7_clinicalplot.pdf',width=12,height=12)
par(mfrow=c(2,2))

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high signal to noise ratio (SNR), and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

Keywords:

* EEG
* VEP
* SSVEP 
* Vision
* Contrast

{{< pagebreak >}}

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965; @schadeOpticalPhotoelectricAnalog1956], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as functional magnetic resonance imaging (fMRI), magnetoencephalography (MEG), and electroencephalography (EEG). Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual visually-evoked potentials (VEPs) reveals multiple generators throughout visual cortex [@Di-Russo2005; @Di-Russo2007]. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies.

Two common stimulus variants involve sinusoidal on-off (or 'appearance/disappearance') flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can excite independent populations of on- and off-cells in the visual system once per cycle. For spatial patterns, the contributions of individual cortical cells to these excitations are thought to sum at the scalp and generate an average of their onset responses on each cycle. For very low spatial frequencies (including zero), resposes at the scalp will be dominated by a single on- or off- cell type on each half cycle.

If the amplitudes and waveforms of responses to the stimulus appearance and disappearnce are not perfectly balanced, a response at the fundamental flicker frequency, known as 1F, and its integer harmonics can be produced. For stimuli of any spatial frequency (including zero), an imbalance can be caused by differences in either the onset- and offset- responses of either on- or off-cells. For example, if only on-sensitive neurons were present (firing to the onset, but not the offset of local increases in luminance) the response to the on/off (or 'appearance/disappearance') presentation of a grating of any spatial frequency below the resolution limit would consist entirely of odd harmonics.

By comparison, counterphase flickering patterns generate two essentially identical transients per cycle and therefore do not produce a response at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker (either on/off or contrast reversing) tends to produce additional spectral harmonics compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while @KaestnerTVST2024 show both different response slopes and thresholds for 1F and 2F components generated by dynamic noise.

SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973; @reganColourCodingPattern1975; @reganEffectStimulusColour1966a; @McKeefry1996; @Di-Russo2001] [see also @Baseler1997].

@@

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@ohzawaContrastGainControl1982; @ohzawaContrastGainControl1985; @bobakCorticalContrastGain1988a; @Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](){#fig-centresurround}
:::
::: {.content-visible unless-format="docx"}
![The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region.](Figures/1_centresurround.pdf){#fig-centresurround}
:::

@@

Although the SSVEP does seem to reflect the activity of relatively well-tuned neuronal populations, changing some aspect of the stimulus may change the nature of those populations. For example, @Campbell1970 noted that densely-sampled measurements of contrast responses can reveal the presence of two qualitatively different types of neurons that exhibit different log-linear contrast functions [see also @Tyler1985; @Nakayama1982; @Norcia1989; @Souza2007]. Modeling such two-limbed contrast response functions using a single sigmoidal function is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an average population response and that the visual system contains many different types of neurons. As a stimulus changes in (for example) contrast, it may selectively activate qualitatively different neuronal populations.

To  understand these processes better, we will show how sinusoidal input signals might be processed by the visual system to produce SSVEPs. @fig-linear illustrates how sine waves of different contrasts are processed in a linear system. The first panel shows the input sine wave, which would be used to modulate stimulus amplitude over time. Notice that there are five peaks in the waveform during the one second sample, so the stimulation frequency is 5Hz (F1). The second panel shows the Fourier transform of the waveform, which contains a substantial peak at this frequency. If we change the stimulus contrast (i.e. the amplitude of the waveform), the amplitude of the F1 component increases linearly with contrast (right panel).

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](){#fig-linear}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](Figures/2_linearsystem.pdf){#fig-linear}
:::

Next we can consider the impact of two types of nonlinearity in processing on the responses.

First, we consider the fact that neurons in the early visual system code either positive or negative contrast. The response of each different type of neuron is an increase in firing rate to the increasing levels of the preferred contrast polarity, and these neurons typically exhibit zero (or baseline) response to the non-preferred polarity. We can model the combined responses of these two cell populations to a time varying contrast signal by full-wave-rectification. The effect of this computation is, effectively, to double the input frequency in the population measurement and to introduce additional higher (even) harmonics due to the discontinuity at the contrast reversals. The effect of full-wave rectification on the SSVEP signal is shown in @fig-fullwaverectification

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE
![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](){#fig-fullwaverectification}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a full-wave rectified sinusoid (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right).](Figures/3_fullwaverectification.pdf){#fig-fullwaverectification}

:::

The contrast response function (CRF) is still linear for this model - rectification has doubled the frequency (and introduced higher harmonics) but the response amplitude measured by the second harmonic is still directly proportional to the input amplitude.

Neurons in the visual system typically do not exhibit this perfect response linearity. Instead, neuronal response are well-modeled by some form of saturating non-linearity: as the contrast increases, the response of the neuron increases but the response increase per unit contrast is reduced at the high end of the contrast range. This saturating nonlinearity is often modeled as a hyperbolic ratio function of the form:

@@

This nonlinearity has clear effects on signal transduction. In @fig-transducer the full-wave rectified waveform (which is combining the response of both 'on' and 'off' neurons) passes through the saturating nonlinearity (left panel of @fig-transducer). The frequency doubling seen in @fig-fullwaverectification is still present but the the contrast response function is now nonlinear (right panel of @fig-transducer).

Interestingly, although the hyperbolic ratio function (@eq-nakarushton) is monotonic, the CRF resulting from measuring the amplitude of the second harmonic (2F) component contains a slight roll-off at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases, and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler1985; @Peirce2007].

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::

To illustrate the effect of contrast gain control [@ohzawaContrastGainControl1982; @Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988; @brownDevelopmentRivalryDichoptic1999; @Candy2001; @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE

![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](){#fig-twoinputs}
:::
::: {.content-visible unless-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](Figures/5_twoinputs.pdf){#fig-twoinputs}
:::

@@

where $C_\mathrm{mask}$ reflects the contrast of the mask component at a frequency distinct from that of the target. The effect of this extra term is to reduce the target response (see right panel of @fig-twoinputs). For a linear contrast axis, the contrast response function becomes shallower, whereas on a logarithmic contrast axis it maintains its steepness and shifts to the right. Suppressive effects of this kind have been obtained using SSVEP with a variety of different types of mask, including orthogonal overlaid masks, surround masks, and dichoptic masks [@Burr1987; @Ross1991; @Candy2001; @Busse2009; @Salelkar2020; @Cunningham2017; for a meta-analysis see @Baker2021].

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the lateral geniculate nucleus (LGN) [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contain information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along the entire processing pathway is challenging, and a full system identification may require the use of multiple different input frequencies [@boydMeasuringVolterraKernels1983], careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing [e.g. @Zemon2006Dynamic] and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, @Tsai2012 demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. More recent work has used this approach to study binocular integration in both normal subjects and amblyopes [@Hou-houchuan2021; @Hou2020]. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very few measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by @bachGreenleeBuhler1988 that examined the effects of adaptation on SSVEP responses, and found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrasts, driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observed in Figure 4.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels (in the linear part of the log-contrast response function prior to saturation), and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis [@Campbell1970] (see @fig-sweepvep). This contrast level has been shown to correspond approximately with psychophysically measured detection thresholds [@campbellVisualEvokedPotential1972; @Norcia1986]. One factor affecting the accuracy of this technique is the presence of noise in the EEG signal which will increase the apparent amplitude of the SSVEP responses across the measurement range. The effect of this noise is particulary apparent for recordings with low SNR or where the slope of the CRF is shallow (and therefore where small changes in the vertical offset of the fit result in large changes in the X-axis intercept). Accurate estimates of thresholds therefore require the experimenter to model this noise and compensate for its effects, as detailed in @Norcia1989.

::: {.content-visible when-format="docx"}
- FIGURE ABOUT HERE
